Lossy compression of DNA sequencing quality data

### 2015-04-21 manuscript/q-compression.tex
Some small edits.

Although we recognize the need for lossless compression for some
purposes and contexts (e.g., archiving, provenance), our perspective
is largely pragmatic with a focus on the use of quality values in
subsequent analyses. From this perspective some loss of information is
deemed acceptable if the inferences from analyses are relatively
unaffected. Here we describe our research investigating lossy
compression of sequence read quality values, specifically those
associated with Illumina instruments, with the objective to provide
some perspective on several strategies rather than to develop a robust
high-quality software for use. Recognizing these properties of
Illumina sequence reads motivates our exploration of three general
classes of lossy compression methods -- binning, modeling, and
profiling -- and consider an exemplar of each class. ~\cite{Canovas:2014fr} and~\cite{janin2013adaptive} evaluated the effects of lossy compression on identifying variants within a data set.  We build on these prior works and access the effects of
% quantifying compression efficiency, we also assess the effects of
quality value information loss resulting from compression on additional
subsequent genomic analyses including read preprocessing (filtering
and trimming), genome assembly, and read mapping.

@@

\textsc{QualComp} is a lossy compression tool that models the quality sequence as a multivariate Gaussian distribution, computing the mean and covariance for each position in the read\citep{Ochoa:2013rt}.
Once the model parameters are calculated, they are stored by the decoder to later reconstruct a \emph{representative} quality value.
\textsc{QualComp} takes as input a user-specified rate (bits per read) and then poses an optimization problem of how to allot these bits for a given position in a quality sequence while minimizing the \textsc{mse}.
The quality sequences can be clustered before hand to produce more accurate models.

@@

As large sets of quality strings show similar trends of quality over
their sequence, it makes sense to identify such common patterns in the
data and use them as reference profiles to approximate individual
quality sequences. Such patterns can be readily determined by
clustering data points (i.e. quality vectors) and using the resulting
cluster centers as representative profiles.

@@

First, the method samples an adjustable number of quality sequences
randomly from the file to be used as a training set. Quality sequences
are represented by vectors containing their \textsc{phred} scores
corresponding to each position along the read. Subsequently, $k$-means
clustering is performed on the training set until convergence. The
obtained cluster centers will be the quality profile prototypes for
the dataset.

Once the $k$ quality profiles are determined, all quality sequences
are passed through the trained $k$-means predictor, with the nearest
quality profile in Euclidean space being assigned to every quality
sequence as their compressed representation.

@@

Although this approach is not able to capture randomly occurring
outlier quality values, it ensures that the overall trends in quality
sequences are retained. Quality profiles capture different overall
qualities and different drop-off positions and gradients. An example
of 128 quality profiles are shown on Figure \ref{fig:profiles_128}.

@@

As the order of the polynomial increases, the bits/base-call increase
and the \textsc{mse} decreases at an exponential rate.  The 7th-degree
polynomial regression has the highest bits/base-call and in the
\textit{Mus musculus} data set, and requires more storage than the
\textsc{ascii} original quality values.  A 7th-degree polynomial
requires storing eight floating point values, resulting in 32 bytes
per quality sequence.  The read length of the \textit{Mus musculus}
data set is only 26, so the 7th-degree polynomial regression is storing
six more bytes than necessary for lossless encoding the quality data.

@@

The 2-bin approach is the only compression method that results in a
higher number of filtered base-pairs across all data sets.  Sickle
uses a sliding window approach to smooth the quality sequences before
it trims the sequence based on a specific quality threshold.  In the
2-bin approach, there is not an even distribution of values per bin.
In other words, \emph{bad} quality values may range from 0-33, whereas
\emph{good} values may only range from 34-40.  Thus, mid-range quality
values that are above the threshold (20 by default) are set below the
quality threshold when compressed, resulting in an increased number of
filtered bases.

The 0-degree polynomial regression results in the highest proportion
of bases kept.  If the average quality value of the sequence is above
the filtering threshold, then no base-pairs are trimmed.  Only
sequences that are comprised of mainly low quality values will be
filtered.

It is important to highlight that the even though a compression method
may result in the same number of filtered base-pairs as the
uncompressed sequences, it does not mean the \emph{same} base-pairs
were filtered.  The 1st-degree and 5th-degree polynomial regression of
the \textit{Rhodobacter sphaeroides} fragment library filters nearly
as many bases as each other.  However, if we examine the specific
sequences filtered, the 5th-degree polynomial regression discards
approximately two thirds less sequences than the 1st-degree polynomial
regression that are kept by the uncompressed sequences (4,640 and
12,066 sequences, respectively) (TODO: move to Supplemental Table
\ref{tab:quality_control}).

@@

The lossy compression methods largely preserve the contiguity found in
the assembly produced using the reads with unmodified quality
sequences. All compression methods other than 0-degree polynomial
regression produce an N50 ranging from 3.17--3.22 Mbps (see
Supplemental Table TODO). Despite the similar contiguity statistics,
the different compression methods vary noticeably in the amount of
\textsc{snp}s. The order of polynomial has an inverse relationship
with the amount of \textsc{snp}s detected. The 2-bin and profile
methods detected the least amount of \textsc{snp}s compared to the
reference genome, outperforming the assembly using the original
quality sequences. A more in-depth evaluation is needed to determine
whether these compression methods are missing actual \textsc{snp}s.

@@

In general, the greater the polynomial order, the better overall
assembly; however, the 5th-degree polynomial regression performs
slightly worse than the 3rd-degree polynomial. The respective ranks in
terms of N50 and relocations are fairly distant, which lowers the
overall ranking of the 5th-degree polynomial slightly below that for
3rd-degree polynomial model. The 1st- and 0-degree polynomial
regression methods perform poor in all metrics except assembled bases.
One explanation is that the high error portions of sequences are being
marked as high quality, so \textsc{allpaths-lg} is unable to trim or
error correct the sequences. Assembled sequences that overlap maybe
unable to align to the errors at the end of the reads, artificially
inflating the assembled genome size.

@@

Certain short read alignment tools use the quality sequence
information when finding potential alignments. Bowtie2 (version 2.2.3)
was used to evaluate the different decompressed \textsc{fastq}
files. Bowtie2 uses quality values written in the \textsc{fastq} files
when mapping reads against a reference genome. The original
uncompressed and decompressed \textsc{fastq} files were mapped with
Bowtie2 against \textit{Rhodobacter sphaeroides} reference genome. The
generated \textsc{sam} file for each compressing approach were
compared with the uncompressed \textsc{sam} file. The total, shared
and unique proportional numbers of mapped reads are calculated with
respect to the uncompressed \textsc{sam} matches as shown in table
\ref{tab:aligner}. Additionally, to monitor the effect of quality
values on mapping in general, Bowtie2 was adjusted so that the maximum
and minimum mismatch penalty were equivalent to maximum and minimum
quality scores (with parameters: --mp 6,6 and --mp 2,2 respectively).

We evaluate the compression methods using two approaches.  In the
first approach, we order the compression methods based on how similar
the alignment results are to the uncompressed sequences, i.e., the
amount of reads aligned by both the uncompressed and compressed
methods plus the amount of reads unaligned by both methods minus the
amount of reads uniquely aligned by the uncompressed and compression
method.  In the second approach, we order the compression methods by
total proportion of aligned reads.

@@

The primary concern of using lossy compression methods is naturally
the extent of information loss, that we quantified by \textsc{mse} in
this study. \textsc{mse} and compressibility provide information in
the theoretical context to the methods, but they are not the end-all
of evaluation criteria. The performance of compressed datasets in
different subsequent analyses and applications are just as
important. Our benchmarks showed that some of the compression methods
with high error rates are still practical for certain kinds of
applications. Many subsequent tools proved to have enough additional
redundancy built-in to handle such loss in information. Passing the
decompressed quality sequences through quality control software shows
that most methods filter nearly as many bases as using original
quality sequences. Assemblers performing sequencing alignment use
percent similarity scores that are typically robust to standard
sequencing errors.

@@

We have simply provided here the initial steps in analyzing the effect
of lossy compression on quality sequences using a single,
high-coverage bacterial data set. More work needs to be done using
additional biological data sets, such as human and mouse, along with
different sequencing technologies. A more direct comparison against
related lossy compression tools, such as
\textsc{SlimGene}~\citep{Kozanitis:2011kl} and
\textsc{\textsc{q}ual\textsc{c}omp}~\citep{Ochoa:2013rt}, needs to be
performed. Additionally, other types of sequencing data can be
analyzed apart from the Illumina data examined here. For example, the
PacBio sequencing instruments produce very long reads (with average
read lengths on the order of 10 kbp), but with the trade-off of having
a high error-rate ($\sim$15\%). Unlike the class of quality values we
have examined here, the distribution of erroneous bases is relatively
uniform~\cite{Ferrarini:2013vf}. The assembly complexity of bacterial
genomes can be greatly simplified, producing near complete genome
assemblies, by utilizing a single run of these long
reads~\citep{Koren:2013ye}. If long read sequencing technologies such
as PacBio become more widely adopted, it would be of huge benefit to
examine the potential of lossy compression algorithms on not only the
quality values, but the biological sequencing data themselves.

>>>>

Although we recognize the need for lossless compression for some
purposes and contexts (e.g., archiving, provenance), our perspective
is largely pragmatic with a focus on the use of quality values in
subsequent analyses. From this perspective some loss of information is
deemed acceptable if the inferences from analyses are relatively
unaffected. Here we describe our research investigating lossy
compression of sequence read quality values, specifically those
associated with Illumina instruments, with the objective to provide
some perspective on several strategies rather than to develop a robust
high-quality software for use. Recognizing these properties of
Illumina sequence reads motivates our exploration of three general
classes of lossy compression methods -- binning, modeling, and
profiling -- and consider an exemplar of each class.~\cite{Canovas:2014fr} and~\cite{janin2013adaptive} evaluated the effects of lossy compression on identifying variants within a data set. We build on these prior works and access the effects of
% quantifying compression efficiency, we also assess the effects of
quality value information loss resulting from compression on additional
subsequent genomic analyses including read preprocessing (filtering
and trimming), genome assembly, and read mapping.

@@

\textsc{QualComp} is a lossy compression tool that models quality values as a multivariate Gaussian distribution, computing the mean and covariance for each position in the read\citep{Ochoa:2013rt}.
Once the model parameters are calculated, they are stored by the decoder to later reconstruct a \emph{representative} quality value.
\textsc{QualComp} takes as input a user-specified rate (bits per read) and then poses an optimization problem of how to allot these bits for a given position while minimizing the \textsc{mse}.
The quality values can be clustered before hand to produce more accurate models.

@@

As large sets of quality strings show similar trends of quality over
their length, it makes sense to identify such common patterns in the
data and use them as reference profiles to approximate individual
sequences of quality values. Such patterns can be readily determined by
clustering data points (i.e. quality vectors) and using the resulting
cluster centers as representative profiles.

@@

First, the method samples an adjustable number of reads
randomly from the file to be used as a training set. Quality values
are represented by vectors containing their \textsc{phred} scores
corresponding to each position along the read. Subsequently, $k$-means
clustering is performed on the training set until convergence. The
obtained cluster centers will be the quality profile prototypes for
the dataset.

Once the $k$ quality profiles are determined, all reads
are passed through the trained $k$-means predictor, with the nearest
quality profile in Euclidean space being assigned to every read
as their compressed representation.

@@

Although this approach is not able to capture randomly occurring
outlier quality values, it ensures that the overall trends in quality
value patterns are retained. Quality profiles capture different overall
qualities and different drop-off positions and gradients. An example
of 128 quality profiles are shown on Figure \ref{fig:profiles_128}.

@@

As the order of the polynomial increases, the bits/base-call increase
and the \textsc{mse} decreases at an exponential rate. The 7th-degree
polynomial regression has the highest bits/base-call and in the
\textit{Mus musculus} data set, and requires more storage than the
\textsc{ascii} original quality values. A 7th-degree polynomial
requires storing eight floating point values, resulting in 32 bytes
per sequence of quality values. The read length of the \textit{Mus musculus}
data set is only 26, so the 7th-degree polynomial regression is storing
six more bytes than necessary for lossless encoding the quality data.

@@

The 2-bin approach is the only compression method that results in a
higher number of filtered base-pairs across all data sets. Sickle
uses a sliding window approach to smooth the read quality values before
it trims the sequence based on a specific quality threshold. In the
2-bin approach, there is not an even distribution of values per bin.
In other words, \emph{bad} quality values may range from 0-33, whereas
\emph{good} values may only range from 34-40. Thus, mid-range quality
values that are above the threshold (20 by default) are set below the
quality threshold when compressed, resulting in an increased number of
filtered bases.

The 0-degree polynomial regression results in the highest proportion
of bases kept. If the mean quality value of the read is above
the filtering threshold, then no base-pairs are trimmed. Only
reads that are comprised of mainly low quality values will be
filtered.

It is important to highlight that the even though a compression method
may result in the same number of filtered base-pairs as the
uncompressed sequences, it does not mean the \emph{same} base-pairs
were filtered. The 1st-degree and 5th-degree polynomial regression of
the \textit{Rhodobacter sphaeroides} fragment library filters nearly
as many bases as each other. However, if we examine the specific
reads filtered, the 5th-degree polynomial regression discards
approximately two thirds less reads than the 1st-degree polynomial
regression that are kept by the uncompressed reads (4,640 and
12,066 reads, respectively) (TODO: move to Supplemental Table
\ref{tab:quality_control}).

@@

The lossy compression methods largely preserve the contiguity found in
the assembly produced using the reads with unmodified quality
values. All compression methods other than 0-degree polynomial
regression produce an N50 ranging from 3.17--3.22 Mbps (see
Supplemental Table TODO). Despite the similar contiguity statistics,
the different compression methods vary noticeably in the amount of
\textsc{snp}s. The order of polynomial has an inverse relationship
with the amount of \textsc{snp}s detected. The 2-bin and profile
methods detected the least amount of \textsc{snp}s compared to the
reference genome, outperforming the assembly using the original
quality values. A more in-depth evaluation is needed to determine
whether these compression methods are missing actual \textsc{snp}s.

@@

In general, the greater the polynomial order, the better overall
assembly; however, the 5th-degree polynomial regression performs
slightly worse than the 3rd-degree polynomial. The respective ranks in
terms of N50 and relocations are fairly distant, which lowers the
overall ranking of the 5th-degree polynomial slightly below that for
3rd-degree polynomial model. The 1st- and 0-degree polynomial
regression methods perform poor in all metrics except assembled bases.
One explanation is that the high error portions of reads are being
marked as high quality, so \textsc{allpaths-lg} is unable to trim or
error correct the sequences. Assembled sequences that overlap maybe
unable to align to the errors at the end of the reads, artificially
inflating the assembled genome size.

@@

Certain short read alignment tools use the quality value
information when finding potential alignments. Bowtie2 (version 2.2.3)
was used to evaluate the different decompressed \textsc{fastq}
files. Bowtie2 uses quality values written in the \textsc{fastq} files
when mapping reads against a reference genome. The original
uncompressed and decompressed \textsc{fastq} files were mapped with
Bowtie2 against \textit{Rhodobacter sphaeroides} reference genome. The
generated \textsc{sam} file for each compressing approach were
compared with the uncompressed \textsc{sam} file. The total, shared
and unique proportional numbers of mapped reads are calculated with
respect to the uncompressed \textsc{sam} matches as shown in table
\ref{tab:aligner}. Additionally, to monitor the effect of quality
values on mapping in general, Bowtie2 was adjusted so that the maximum
and minimum mismatch penalty were equivalent to maximum and minimum
quality scores (with parameters: --mp 6,6 and --mp 2,2 respectively).

We evaluate the compression methods using two approaches. In the
first approach, we order the compression methods based on how similar
the alignment results are using the original uncompressed quality values, i.e., the
amount of reads aligned by both the uncompressed and compressed
methods plus the amount of reads unaligned by both methods minus the
amount of reads uniquely aligned by the uncompressed and compression
method. In the second approach, we order the compression methods by
total proportion of aligned reads.

@@

The primary concern of using lossy compression methods is naturally
the extent of information loss, that we quantified by \textsc{mse} in
this study. \textsc{mse} and compressibility provide information in
the theoretical context to the methods, but they are not the end-all
of evaluation criteria. The performance of compressed datasets in
different subsequent analyses and applications are just as
important. Our benchmarks showed that some of the compression methods
with high error rates are still practical for certain kinds of
applications. Many subsequent tools proved to have enough additional
redundancy built-in to handle such loss in information. Passing the
decompressed quality values through quality control software shows
that most methods filter nearly as many bases as using original
quality sequences. Assemblers performing sequence alignment use
percent similarity scores that are typically robust to standard
sequencing errors.

@@

We have simply provided here the initial steps in analyzing the effect
of lossy compression on quality values using a single,
high-coverage bacterial data set. More work needs to be done using
additional biological data sets, such as human and mouse, along with
different sequencing technologies. A more direct comparison against
related lossy compression tools, such as
\textsc{SlimGene}~\citep{Kozanitis:2011kl} and
\textsc{\textsc{q}ual\textsc{c}omp}~\citep{Ochoa:2013rt}, needs to be
performed. Additionally, other types of sequencing data can be
analyzed apart from the Illumina data examined here. For example, the
PacBio sequencing instruments produce very long reads (with average
read lengths on the order of 10 kbp), but with the trade-off of having
a high error-rate ($\sim$15\%). Unlike the class of quality values we
have examined here, the distribution of erroneous bases is relatively
uniform~\cite{Ferrarini:2013vf}. The assembly complexity of bacterial
genomes can be greatly simplified, producing near complete genome
assemblies, by utilizing a single run of these long
reads~\citep{Koren:2013ye}. If long read sequencing technologies such
as PacBio become more widely adopted, it would be of huge benefit to
examine the potential of lossy compression algorithms on not only the
quality values, but the biological sequencing data themselves.
